\chapter{Biaflows}
\label{app:biaflows}

\paragraph{Abstract.} Image analysis is key to extracting quantitative information from scientific microscopy images, but the methods involved are now often so refined that they can no longer be unambiguously described by written protocols. We introduce BIAFLOWS, an open-source web tool enabling to reproducibly deploy and benchmark bioimage analysis workflows coming from any software ecosystem. A curated instance of BIAFLOWS populated with 34 image analysis workflows and 15 microscopy image datasets recapitulating common bioimage analysis problems is available online. The workflows can be launched and assessed remotely by comparing their performance visually and according to standard benchmark metrics. We illustrated these features by comparing seven nuclei segmentation workflows, including deep-learning methods. BIAFLOWS enables to benchmark and share bioimage analysis workflows, hence safeguarding research results and promoting high-quality standards in image analysis. The platform is thoroughly documented and ready to gather annotated microscopy datasets and workflows contributed by the bioimaging community.

\paragraph{Publication.} Ulysse Rubens$^\ast$, Romain Mormont$^\ast$, et al. ``BIAFLOWS: A collaborative framework to reproducibly deploy and benchmark bioimage analysis workflows''. In: \textit{Patterns}  1.3 (2020), p. 100040 ($\ast$These authors contributed equally).

\paragraph{Article.} \url{https://www.cell.com/patterns/fulltext/S2666-3899%2820%2930045-3}

\paragraph{Code.} \url{https://github.com/Neubias-WG5}

\paragraph{Documentation.} \url{http://biaflows-doc.neubias.org/}
